Is high social class always beneficial for survival? Northern Sweden 1801--2013

### 2018-02-13 article3b.Rmd
Some corrections

The results showing a small or even reversed socio-economic gradient in the past and its late development are in accordance with other studies [@bevp11], for example @bedr11 in a different setting, a couple of Scanian parishes in southern Sweden, albeit with somewhat different social categories. They find the first indications of an emerging social gradient to be a fairly recent phenomenon, in their case as late as the 1950s. Compared to previous studies, the present one shows a much clearer survival disadvantage of the elite, and not only the lack of social gradient. There are in fact some studies indicating a similar pattern. A study of mortality and social class in Sweden during the period 1960--1968 found that self-employed men and non-manual male workers in fact had higher mortality than manual workers in total as well as in cardiovascular mortality [@vano89]. A recent study by @erdr17 on the social pattern of mortality, based on 1880--1910 census data linked to death registers, shows that the unexpected pattern found in the present study in not unique, but rather seems to have characterised all parts of the country. These findings are contrary to what is usually assumed and what other studies have shown. The studies usually referred to when discussing the historically large social inequalities in health are mainly based on data from England and Wales [@ant67; @wowi95; @pamuk85]. We know less about how general these results are, but apparently the Swedish context differed from the English and Welsh ones.

@@

We believe our results have implications on several issues concerning our understanding of the demographic and epidemiologic development. One of the central theories has been presented by @mck76, who argues that improved nutrition has been the main determinant of the increase in life expectancy in recent centuries. This ought to imply that people with less adequate resources are the most vulnerable in the earlier phases of the mortality transition. The results presented here prove that the most wealthy and privileged groups were not able to avoid the most common health risks in historical societies. This should make us cautious of accepting nutritional improvements as the most important explanation for the mortality transition. While the McKeown thesis has been refuted by several scholars [@szreter88, @colgrove02], it has played and is still playing a central role within epidemiology and social medicine.

@@

We must thus consider other and potentially more important explanations than the pure advantage of being in a high social position, reflecting that the embodiment can be materialized in different ways. In the following we discuss three possible pathways -- the life course perspective, social stress and class- and gender-based lifestyles. Concerning the life course perspective, exposure to diseases and other negative experiences in childhood can influence one's health later in life. Up until the early 20th century, being exposed to an urban environment had far higher health consequences than  living in rural Sweden. We control for the type of environment people live in, but we do not have full information about previous residences. There is a possibility that the elite group to a greater extent had spent their childhood in the more harmful urban environments, and that this might have influenced their health and survival in adulthood. It seems improbable, however, that differences in residential backgrounds could explain most of the observed pattern. The second pathway relates to the health risks of belonging to a certain social class. We usually assume that health risks are higher in social classes with less access to resources. It is possible, however, that men in high positions had a stressful life, for example entrepreneurs facing the risk of losing an advantageous position. Business often failed which meant a substantial loss of prestige and social network. Suicides certainly occurred among the elite, but it is difficult to verify the extent of social stress in this and other social classes.

Social stress can also be expressed in unhealthy lifestyles, the third pathway we discuss. Harmful behaviours can be connected to certain expectations associated with being a man of high status. Our results regarding the gender aspect are important in this respect. As several researchers have pointed out, demographic research has often ignored power relations and different roles within families [@watkins93; @mack95; @jans07]. The gender difference in mortality may be caused by discrimination of the female party in the household. Historical Sweden was strongly gender-segregated, discriminating against women and giving them less space for action. In our study, we do not find the lack of power to have negative health consequences for women. Within the same households, the effect of social class leads to different results, high social position being beneficial for women but not for men. Instead, we might consider the role of behaviours based on distinct gendered and class expectations. Having a high social status is sometimes expressed in attitudes, norms and way of living that are potentially harmful to one's health. @edli11, finding similar results in another environment, suggest that an explanation for the large social health inequalities in present-day society is that high status is expressed through health, being fit and physically active, while in many historical contexts higher social status (for men) was expressed in low physical activity, obsessive eating, drinking and smoking and being more inclined to take risks.  Alcohol has been suggested to be part of the large gender differences in the 19th century [@wil99; @edv92], and a more risk-taking lifestyle may be an important component of male mortality [@cour00; @rasp06; @jaadla17]. Different historical contexts promote different ways to spend economic resources.

@@

Overweight, obesity and physical inactivity were probably more common in higher social classes. @vano89 find a higher total as well as cardiovascular mortality among self-employed men and non-manual male workers, which they suggest can be explained by the higher tobacco consumption among men in these groups. The social pattern among women was the reverse, however, in line with the present study. There is no solid information on smoking in different social classes before the 1960s, but there are international studies indicating a higher prevalence and consumption of tobacco in higher classes during the early phases of the tobacco epidemic, for example illustrated by the fact that smoking was initially more prevalent among physicians. There was also a clear association between higher incomes and tobacco consumption [@loco94].

The same lack of statistical evidence characterizes knowledge about alcohol consumption in different social classes in Sweden before 1960, but with some exceptions. Access to alcohol was much easier for the wealthy during the period of strong alcohol restriction in 1915-1955 and the purchase of alcoholic beverages was substantially higher among those with the highest incomes. There was also an overrepresentation of alcohol-related deaths in the higher social classes in the 1960s: @noro98 suggest that during the last decades of the 20th century, there was an equalization and even a reversal of alcohol consumption between social strata, with higher levels among manual labourers but also with an increasingly skewed consumption distribution, indicating more high consumers in this category. A similar reversal has been observed in England and Wales during the 20th century when it comes to cirrhosis mortality, possibly reflecting a change in consumption patterns. Mortality from this disease was substantially higher in the managerial and professional classes in 1921, being reversed in 1991 with the highest mortality in the lowest class [@cropr10]. It could be discussed whether these indications of changes in health-related behaviour are sufficiently significant to explain the present findings; we believe they are certainly at least an important part of the story.

@@

An important question is whether our results can be generalized to other historical contexts. We have studied environments in northern Sweden with no large metropolitan areas. Perhaps Sweden or this part of Sweden is a special case. The involvement of government at both the national and local levels may have been stronger, thus mitigating socio-economic disadvantages, but this is doubtful concerning the 19th century. The level of income inequality was certainly lower in our regions than in other parts of the country but it was still substantially higher in 1930 than towards the end of the 20th century. Compared with many other countries, the level of mortality has been fairly low. The question of generalizability is relevant, which can hopefully be addressed in other studies, preferably with a comparative approach.

>>>>

The results showing a small or even reversed socio-economic gradient in the past and its late development are in accordance with other studies [@bevp11], for example by @bedr11 in a different setting, a couple of Scanian parishes in southern Sweden, albeit with somewhat different social categories. They find the first indications of an emerging social gradient to be a fairly recent phenomenon, in their case as late as the 1950s. Compared to previous studies, the present one shows a much clearer survival disadvantage of the elite, and not only the lack of social gradient. There are in fact some studies indicating a similar pattern. A study of mortality and social class in Sweden during the period 1960--1968 found that self-employed men and non-manual male workers in fact had higher mortality than manual workers in total as well as in cardiovascular mortality [@vano89]. A recent study by @erdr17 on the social pattern of mortality, based on 1880--1910 census data linked to death registers, shows that the unexpected pattern found in the present study in not unique, but rather seems to have characterised all parts of the country. These findings are contrary to what is usually assumed and what other studies have shown. The studies usually referred to when discussing the historically large social inequalities in health are mainly based on data from England and Wales [@ant67; @wowi95; @pamuk85]. We know less about how general these results are, but apparently the Swedish context differed from the English and Welsh ones.

@@

We believe our results have implications on several issues concerning our understanding of the demographic and epidemiologic development. One of the central theories has been presented by @mck76, who argues that improved nutrition has been the main determinant of the increase in life expectancy in recent centuries. This ought to imply that people with less adequate resources are the most vulnerable in the earlier phases of the mortality transition. The results presented here prove that the most wealthy and privileged groups were not able to avoid the most common health risks in historical societies. This should make us cautious of accepting nutritional improvements as the most important explanation for the mortality transition. While the strong version of the McKeown thesis has been refuted by several scholars [@szreter88, @colgrove02], it continues to play a central role within epidemiology and social medicine.

@@

We must thus consider other and potentially more important explanations than the pure advantage of being in a high social position. In the following we discuss three possible pathways -- the life course perspective, social stress and class- and gender-based lifestyles. Concerning the life course perspective, exposure to diseases and other negative experiences in childhood can influence one's health later in life. Up until the early 20th century, being exposed to an urban environment had far higher health consequences than  living in rural Sweden. We control for the type of environment people live in, but we do not have full information about previous residences. There is a possibility that the elite group to a greater extent had spent their childhood in the more harmful urban environments, and that this might have influenced their health and survival in adulthood. It seems improbable, however, that differences in residential backgrounds could explain most of the observed pattern. The second pathway relates to the health risks of belonging to a certain social class. We usually assume that health risks are higher in social classes with less access to resources. It is possible, however, that men in high positions had a stressful life, for example entrepreneurs facing the risk of losing an advantageous position. Business often failed which meant a substantial loss of prestige and social network. Suicides certainly occurred among the elite, but it is difficult to verify the extent of social stress in this and other social classes.

Social stress can also be expressed in unhealthy lifestyles -- the third pathway. Harmful behaviours can be connected to certain expectations associated with being a man of high status. Our results regarding the gender aspect are important in this respect. As several researchers have pointed out, demographic research has often ignored power relations and different roles within families [@watkins93; @mack95; @jans07]. The gender difference in mortality may be caused by discrimination of the female party in the household. Historical Sweden was strongly gender-segregated, discriminating against women and giving them less space for action. In our study, we do not find the lack of power to have negative health consequences for women. Within the same households, the effect of social class leads to different results, high social position being beneficial for women but not for men. Instead, we might consider the role of behaviours based on distinct gendered and class expectations. Having a high social status is sometimes expressed in attitudes, norms and way of living that are potentially harmful to one's health. @edli11, finding similar results in another environment, suggest that an explanation for the large social health inequalities in present-day society is that high status is expressed through health, being fit and physically active, while in many historical contexts higher social status (for men) was expressed in low physical activity, obsessive eating, drinking and smoking and being more inclined to take risks.  Alcohol has been suggested to be part of the large gender differences in the 19th century [@wil99; @edv92], and a more risk-taking lifestyle may be an important component of male mortality [@cour00; @rasp06; @jaadla17]. Different historical contexts promote different ways to spend economic resources.

@@

Overweight, obesity and physical inactivity were probably more common in higher social classes. @vano89 find a higher total as well as cardiovascular mortality among self-employed men and non-manual male workers in Sweden during the 1960s, which they suggest can be explained by the higher tobacco consumption among men in these groups. The social pattern among women was the reverse, however, in line with the present study. There is no solid information on smoking in different social classes before the 1960s, but there are international studies indicating a higher prevalence and consumption of tobacco in higher classes during the early phases of the tobacco epidemic, for example illustrated by the fact that smoking was initially more prevalent among physicians. There was also a clear association between higher incomes and tobacco consumption [@loco94].

The same lack of statistical evidence characterizes knowledge about alcohol consumption in different social classes in Sweden before 1960, but with some exceptions. Access to alcohol was much easier for the wealthy during the period of strong alcohol restriction in 1915-1955 and the purchase of alcoholic beverages was substantially higher among those with the highest incomes. There was also an overrepresentation of alcohol-related deaths in the higher social classes in the 1960s: @noro98 suggest that during the last decades of the 20th century, there was an equalization and even a reversal of alcohol consumption between social strata, with higher levels among manual labourers but also with an increasingly skewed consumption distribution, indicating more high consumers in this category. A similar reversal has been observed in England and Wales during the 20th century when it comes to cirrhosis mortality, possibly reflecting a change in consumption patterns. Mortality from this disease was substantially higher in the managerial and professional classes in 1921, being reversed in 1991 with the highest mortality in the lowest class [@cropr10]. Whether these indications of changes in health-related behaviour are sufficiently significant to explain the present findings can be discussed; we believe they certainly are at least an important part of the story.

@@

An important question is whether our results can be generalized to other historical contexts. We have studied environments in northern Sweden with no large metropolitan areas. Perhaps Sweden or this part of Sweden is a special case. The involvement of government at both the national and local levels may have been stronger, thus mitigating socio-economic disadvantages, but this is doubtful concerning the 19th century. The level of income inequality was certainly lower in our regions than in other parts of the country but it was still substantially higher in 1930 than towards the end of the 20th century. Compared with many other countries, the mortality level has been fairly low. The question of generalizability is relevant, which can hopefully be addressed in other studies, preferably with a comparative approach.
